---
title: "Figure 4"
author: 
  - "Matt Paul"
  - "Doug Barrows"
  - "Bioinformatics Resource Center"
  - "Yakshi Dabas"
  - "Allis Lab"
  - "Rockefeller University"
date: '2023-04-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Figure 4
This is a collection of code that was used to generate the plots in Figure 4. 

## Figure 4

### Figure 4A + 4B
- GO term analysis

```{r}
library(magrittr)
library(tibble)
library(ggplot2)
library(dplyr)
library(GO.db)
library(topGO)
library(org.Mm.eg.db)
library(biomaRt)
library(annotate)

A <- DE_table_WTR26 %>% as_tibble(rownames = 'A') %>% filter(padj<0.05) %>% pull(A)
D <- DE_table_WTR2 %>% as_tibble(rownames = 'D') %>% filter(padj<0.05) %>% pull(D)
ENTREZID<-intersect(A,D)

R2CvsR26C <- cbind(ENTREZID, DE_table_WTR2[rownames(DE_table_WTR2) %in%ENTREZID,]$log2FoldChange, DE_table_WTR26[rownames(DE_table_WTR26) %in%ENTREZID,]$log2FoldChange)

R2CvsR26C %<>% as_tibble() %>% as_tibble() %>% dplyr::rename(log2FC_R2C=V2) %>% dplyr::rename(log2FC_R26C=V3) %>% filter(log2FC_R2C>0&log2FC_R26C>0 |  log2FC_R2C<0&log2FC_R26C<0)

db_mm= useMart('ENSEMBL_MART_ENSEMBL',dataset="mmusculus_gene_ensembl")
go_ids= getBM(attributes=c('go_id', "entrezgene_id", 'namespace_1003'), mart=db_mm)
gene_2_GO=unstack(go_ids[,c(1,2)])
all_genes<-names(gene_2_GO)

doGO<-function(IDs){
# remove any candidate genes without GO annotation
keep <- IDs %in% all_genes
keep <- which(keep==TRUE)
test_genes <- IDs[keep]

if(length(test_genes)!=0){
# make named factor showing which genes are of interest
geneList <- factor(as.integer(all_genes %in% IDs))
names(geneList) <- all_genes

GOdata<-new("topGOdata",ontology = "BP",allGenes = geneList,nodeSize = 4, annot= annFUN.gene2GO, gene2GO =gene_2_GO)

resultfisher <- runTest(GOdata, "classic", "fisher")
resultfisher.weight <- runTest(GOdata, "weight", "fisher")

      
allRes <- GenTable(GOdata, resultfisher,resultfisher.weight,topNodes=length(GOdata@graph@nodes))
}else{
 allRes<-data.frame("No Overlap Genes")
}

return(allRes)}

R2CvsR26C_mouse_go<-doGO(R2CvsR26C %>% pull(ENTREZID))


```

```{py}
 # Create a csv file from TOPGO analysis with first column depicting GO Terms, 2nd column depicting -log10(result2), 3rd column depicting Significant/Expected
#Import libraries
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.colors as colors
import matplotlib.cbook as cbook
from matplotlib.cm import ScalarMappable
import matplotlib
from mpl_toolkits.axes_grid1 import make_axes_locatable

#Set plotting parameters
matplotlib.rcParams['mathtext.fontset'] = 'custom'
matplotlib.rcParams['mathtext.it'] = 'Arial:italic'
matplotlib.rcParams['mathtext.rm'] = 'Arial'
matplotlib.rcParams['font.family'] = 'Arial'
matplotlib.rcParams['axes.linewidth'] = 4
matplotlib.rcParams['xtick.major.size'] = 10
matplotlib.rcParams['xtick.major.width'] = 4
matplotlib.rcParams['ytick.major.size'] = 10
matplotlib.rcParams['ytick.major.width'] = 4
matplotlib.rcParams.update({'font.size': 30})
matplotlib.rc('font', weight=5)
lw_param = 3
colors = ['tab:blue','tab:green','tab:purple','tab:gray','tab:olive'
          ,'tab:orange','tab:red','tab:brown','tab:pink','tab:cyan'] 
fig, (ax1) = plt.subplots(figsize=(15, 10),nrows=1, ncols=1)

# Read csv data
file_name = "R2CvsR26C_mouse_same_direction_GO.csv"
data = np.loadtxt(file_name,delimiter = ',',dtype = str)
name = data[:,0]
Sig_Exp = np.float64(data[:,2])
p_val = np.float64(data[:,1])
Z= Sig_Exp
Y= name
X= p_val


data_color = [x / max(Z) for x in Z]

# Generate plot
plt_ax = ax1.scatter(X, Y, s= 500, c=Z, cmap='winter')

ax1.set_xlabel("-Log$_{10}$ [Elim statistic]")
ax1.set_ylabel(" GO terms")

cbar1 = plt.colorbar(plt_ax,ax=ax1)
cbar1.ax.set_title("Sig/Exp")  

red_list = [6,8,9,10,13,15,17,18]

for itr, t in enumerate(ax1.yaxis.get_ticklabels()):
    print(itr)
    if itr in red_list:
        t.set_color('red') 

plt.savefig('R2CvsR26C_mouse_same_direction_GO.pdf',dpi =900, bbox_inches = "tight")

```

### Figure 4C

- Violin plots of RNAseq at significant C+R peaks

```{r}
dge_temp <- rio::import("GEO/final files/ChIP_R26C_minus_WT_annotated.xlsx")
rna_res <- DE_table_WTR26
```

```{r}
is.na(dge_temp$padj)[is.na(dge_temp$padj) ]<-1
all_temp <- rna_res %>% as_tibble(rownames = "ENTREZ") %>% full_join(dge_temp, by=c("ENTREZ" = "geneId")) %>% dplyr::mutate(class=ifelse(!(padj.y<0.001), "NS",ifelse(log2FoldChange.y>0,"Up", "Down"))) %>% dplyr::filter(distanceToTSS==0)
all_temp <- all_temp[!is.na(all_temp$class), ]
  
all_temp %>% mutate(class=factor(class, levels=c("Up", "Down", "NS"))) %>% ggplot( aes(x=log2FoldChange.x, y=class)) +geom_violin() +coord_flip() +ggtitle(paste0("L2fc gene expression at genes with H3K27me3 SEACR peaks \nthat significantly change (padj<0.001), comparing WT to mutant R26C")) + xlab("log2FC(Mutant/WT)")+ theme_bw() + geom_signif(test="wilcox.test", comparisons = combn(unique(all_temp$class),2, simplify = F),step_increase = 0.2)

```

### Figure 4D
- Motif analysis at R26C

```{R}

install_CondaTools("meme", "meme", pathToMiniConda = "../mini")
meme_path_id="../mini/envs/meme/bin/"

download.file("https://meme-suite.org/meme/meme-software/Databases/motifs/motif_databases.12.24.tgz", "motif_databases.12.24.tgz")

library(universalmotif)
library(magrittr)
meme_path_id="../mini/envs/meme/bin/"

meme_db <- read_meme("motif_databases/CIS-BP_2.00/Mus_musculus.meme") %>% to_df()

```

```{r}
library(GenomicRanges)
library(org.Mm.eg.db)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(dplyr)
library(BSgenome.Mmusculus.UCSC.mm10)
library(memes)

DE_table <- rio::import("GEO/final files/ChIP_R26C_minus_WT_annotated.xlsx")
dge_temp <- DE_table

dge_gr <- GRanges(dge_temp$seqnames, IRanges(dge_temp$start, dge_temp$end ))

anno_1 <- ChIPseeker::annotatePeak(dge_gr ,TxDb = TxDb.Mmusculus.UCSC.mm10.knownGene, annoDb = "org.Mm.eg.db", verbose=F)
dge_temp <- cbind(as.data.frame(anno_1),dge_temp)
is.na(dge_temp$padj)[is.na(dge_temp$padj) ] <- 1

colnames(dge_temp)[1:5] <-  paste0(colnames(dge_temp)[1:5],"_2")
all_temp <- dge_temp %>% dplyr::mutate(class=ifelse(!(padj<0.001), "NS",ifelse(log2FoldChange>0,"Up", "Down"))) %>% dplyr::filter(distanceToTSS==0)

all_temp <- all_temp[!is.na(all_temp$class), ]
all_temp %<>% mutate(class=factor(class, levels=c("Up", "Down", "NS"))) 
all_temp <- all_temp[!is.na(all_temp$log2FoldChange),]

all_genes <- genes(TxDb.Mmusculus.UCSC.mm10.knownGene)
promoter_ranges <- promoters(all_genes,  upstream=1000, downstream=200)
promoter_ranges_alldownpeaks <- promoter_ranges[promoter_ranges$gene_id %in% all_temp$geneId[all_temp$class=="Down"]]

peaksSequences_alldownpeaks <- getSeq(BSgenome.Mmusculus.UCSC.mm10, promoter_ranges_alldownpeaks)
writeXStringSet(peaksSequences_alldownpeaks, file = "prom_alldownpeaks.fa")
peaksSequences_allpeaks <- getSeq(BSgenome.Mmusculus.UCSC.mm10, promoter_ranges)
writeXStringSet(peaksSequences_allpeaks, file = "prom_allpeaks.fa")

ame_by_binding_shuff <- runAme("prom_alldownpeaks.fa", meme_path = meme_path_id, database = "motif_databases/CIS-BP_2.00/Mus_musculus.meme", silent = FALSE)

ame_by_binding_background <- runAme("prom_alldownpeaks.fa", control ="prom_allpeaks.fa", meme_path = meme_path_id, database = "motif_databases/CIS-BP_2.00/Mus_musculus.meme", silent = FALSE)

results <- list(ame_by_binding_shuff,
                ame_by_binding_background)

names(results) <-c("shuffled_background", "promoter_background")

rio::export(results, "CISBP_2.00_motif_analysis.xlsx")
```

```{r}
library(magrittr)
library(stringr)

library(motifStack)
motifs_hit <- rio::import("CISBP_2.00_motif_analysis.xlsx")
motifs <- importMatrix("motif_databases/CIS-BP_2.00/Mus_musculus.meme")

my_names <- names(motifs) %>% str_split("2.00", simplify = T) %>% .[,1] %>% paste0(.,"2.00")

motif_sub <- motifs[my_names %in% motifs_hit$motif_id]

```

```{r motifs, fig.width=11, fig.height=10, dpi=90, fig.path='plots/', dev=c('png', 'pdf'), eval=F}

sapply(1:length(motif_sub), function(x){
  
  a <- plot(motif_sub[[x]])
  print(a)
})

```

### Figure 4E
- Gene set enrichment analysis of R26vsWT. Pathways are from MSigDB Hallmarks

```{r}

rnk_df <- DE_table_WTR26 %>% na.omit %>%
  arrange(desc(stat)) %>%
  dplyr::select(1, stat)
rnk_vector <- rnk_df[,2]
names(rnk_vector) <- rnk_df[,1]

over_rep_active_categories <- NULL
gsea_active_categories <- c('MSigDB hallmark' = "msig.H")

category_list <- list()

if (!is.null(gsea_active_categories)){
  for(i in seq_along(gsea_active_categories)){
    gsea_term_tmp <- gsea_active_categories[i]
    category_list[["gsea"]][[names(gsea_term_tmp)]] <- str_split(gsea_term_tmp, pattern = "\\.")
  }
}

over_rep_output_list <- list()
gsea_output <- list()
GO_to_gene_lists <- list()

i=1
x=1

active_category <- names(category_list[[i]])[x] 
category <- category_list[[i]][[x]][[1]][[2]]
      if (length(category_list[[i]][[x]][[1]]) == 2){
        subcategory <- NULL
      }else if(length(category_list[[i]][[x]][[1]]) == 3){
        subcategory <- category_list[[i]][[x]][[1]][[3]]
      }
      

msig_t2g <- msigdbr(species = "Mus musculus", 
                          category = category, 
                          subcategory = subcategory) %>% 
        dplyr::select(gs_name, entrez_gene)
      
      GO_to_gene_lists[[active_category]] <- split(msig_t2g, msig_t2g$gs_name) %>%
        lapply(as.data.frame) %>%
        lapply(function(x) x[,2])
      
        set.seed(0)
        gsea_output[[active_category]] <- GSEA(rnk_vector, 
                                               TERM2GENE = msig_t2g, 
                                               seed = TRUE, 
                                               pvalueCutoff = 1,
                                               eps = 1e-100)



        
gsea_output_df <- list()
if (!is.null(gsea_active_categories)){
  for(i in seq_along(gsea_active_categories)){
    gsea_term_tmp <- names(gsea_active_categories)[i]
    gsea_output_df[[gsea_term_tmp]] <- data.frame(gsea_output[[gsea_term_tmp]])
    genes_in_cat <- gsea_output_df[[gsea_term_tmp]]$core_enrichment
    genes_in_cat <- str_split(genes_in_cat, pattern = "\\/")
    symbol_genes_in_cat <- sapply(1:length(genes_in_cat), function(x){ AnnotationDbi::select(x=org.Mm.eg.db,keys=genes_in_cat[[x]],columns="SYMBOL",keytype="ENTREZID")[,2] })
    
    # this last column is a list, we will turn into a character vector
    symbol_genes_in_cat <- lapply(symbol_genes_in_cat, paste, collapse = ",")
    core_enrichment_symbol <- do.call(rbind, symbol_genes_in_cat)
    gsea_output_df[[gsea_term_tmp]] <- cbind(gsea_output_df[[gsea_term_tmp]], 
                                             core_enrichment_symbol)
    
  }
}
```

```{r}
gseaplot2(gsea_output[[active_category]], geneSetID = "HALLMARK_ANGIOGENESIS")
gseaplot2(gsea_output[[active_category]], geneSetID = "HALLMARK_MYC_TARGETS_V1")
gseaplot2(gsea_output[[active_category]], geneSetID = "HALLMARK_HYPOXIA")
gseaplot2(gsea_output[[active_category]], geneSetID = "HALLMARK_MYOGENESIS")

```

## Figure S4
### Figure S4A
- Volcano plots of C+R

```{r}
DE_table_WTR2_CR <- read.table("GEO/final files/CR_H3K27me3atH3K27me3_R2C_minus_WT_DEG.tsv", header = T)
DE_table_WTR26_CR <- read.table("GEO/final files/CR_H3K27me3atH3K27me3_R26C_minus_WT_DEG.tsv", header = T)

EnhancedVolcano(DE_table_WTR2_CR,
    labSize=2,
    title = paste("K27me3 signal at K27me3 SEACR peaks \n in WT vs R2C"),
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.05,
    subtitle ="",
    ylab = bquote(~-Log[10]~italic(padj)),
    legendLabels = c('NS', expression(Log[2]~FC),
    'padj', expression(padj~and~log[2]~FC)))

EnhancedVolcano(DE_table_WTR26_CR,
    labSize=2,
    title = paste("K27me3 signal at K27me3 SEACR peaks \n in WT vs R62C"),
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.05,
    subtitle ="",
    ylab = bquote(~-Log[10]~italic(padj)),
    legendLabels = c('NS', expression(Log[2]~FC),
    'padj', expression(padj~and~log[2]~FC)))

```

### Figure S4B and S4C
- Region plots of C+R

```{r}

chip_bw <- dir("/GEO/final files/CR/H3K27me3/mean/", pattern=".bw", full.names = T)

bed <- paste0("/GEO/final files/CR/H3K27me3/K27_consensus_peaks.bed")

chipProfile_peaks <- BamBigwig_to_chipProfile(signalFiles = chip_bw,
                         testRanges = bed,
                         format = "bigwig",
                         style = "point", 
                         distanceAround=10000,
                         bin_size = 200)

proplyr_object_peaks <- as_profileplyr(chipProfile_peaks)

row.names(sampleData(proplyr_object_peaks)) <- chip_bw %>% basename()

rowData(proplyr_object_peaks )$sgGroup<-'K27me3 at center of K27me3 SEACR peak +/- 10kb'

generateEnrichedHeatmap(proplyr_object_peaks, matrices_color = c("blue", "yellow", "red"))
```

```{r}

chip_bw <- dir("/GEO/final files/CR/H3K27me3/singlerep/", pattern=".bw", full.names = T)

bed <- paste0("/GEO/final files/CR/H3K27me3/K27_consensus_peaks.bed")

chipProfile_peaks <- BamBigwig_to_chipProfile(signalFiles = chip_bw,
                         testRanges = bed,
                         format = "bigwig",
                         style = "point", 
                         distanceAround=10000,
                         bin_size = 200)

proplyr_object_peaks <- as_profileplyr(chipProfile_peaks)

row.names(sampleData(proplyr_object_peaks)) <- chip_bw %>% basename()

rowData(proplyr_object_peaks )$sgGroup<-'K27me3 at center of K27me3 SEACR peak +/- 10kb'

generateEnrichedHeatmap(proplyr_object_peaks, matrices_color = c("blue", "yellow", "red"))

```

### Figure S4D

```{r}
#10T
library(GenomicRanges)
library(magrittr)
library(ggplot2)
library(SummarizedExperiment)
library(tidyr)

load("K27_counts.RData") 
load("K27_consensus_peaks.RData") 

R2C_peaks <- apply(mcols(allPeaksSet_consensus)[,c("R2C_R1","R2C_R2","R2C_R3")],1, all)
R26C_peaks <- apply(mcols(allPeaksSet_consensus)[,c("R26C_R1","R26C_R2","R26C_R3")],1, all)

count_sums <- assay(this_Counts) %>% apply(2, sum)
this_Counts_cpm <- sapply(1:length(count_sums), function(x){
  count_out <- assay(this_Counts)[,x]/count_sums[x]
  count_out <- (count_out)*1000000
  return(count_out)
})
colnames(this_Counts_cpm) <- colnames(this_Counts)

this_Counts_R2C <- this_Counts_cpm[R2C_peaks,] %>% .[,c(1:6)]
this_Counts_R26C <- this_Counts_cpm[R26C_peaks,] %>% .[,c(1:3,7:9)]

colnames(this_Counts_R2C) <- c("R2C_R1","R2C_R2","R2C_R3", "WT_R1","WT_R2","WT_R3")
colnames(this_Counts_R26C) <- c("R26C_R1","R26C_R2","R26C_R3", "WT_R1","WT_R2","WT_R3")

this_Counts_R2C_df <- rbind(this_Counts_R2C[,c(1,4)],
                            this_Counts_R2C[,c(2,5)],
                            this_Counts_R2C[,c(3,6)])

this_Counts_R26C_df <- rbind(this_Counts_R26C[,c(1,4)],
                            this_Counts_R26C[,c(2,5)],
                            this_Counts_R26C[,c(3,6)])

colnames(this_Counts_R2C_df) <- c("WT","R2C")
colnames(this_Counts_R26C_df) <- c("WT","R26C")

```

```{r}

ggplot(as.data.frame(this_Counts_R2C_df), aes(x = WT, y = R2C, alpha=0.4)) + geom_point() + geom_abline(intercept = 0, slope = 1, color = "red") + theme_bw() + labs(title = "R2C - 10T - K27me3 - Counts", x = "WT", y = "R2C") + 
  scale_y_continuous(trans='log2', limits = c(1, 16500)) +
  scale_x_continuous(trans='log2', limits = c(1, 16500))

```

```{r}

ggplot(as.data.frame(this_Counts_R26C_df), aes(x = WT, y = R2C, alpha=0.4)) + geom_point() + geom_abline(intercept = 0, slope = 1, color = "red") + theme_bw() + labs(title = "R2C - 10T - K27me3 - Counts", x = "WT", y = "R2C") + 
  scale_y_continuous(trans='log2', limits = c(1, 16500)) +
  scale_x_continuous(trans='log2', limits = c(1, 16500))

```

### Figure S4E

```{r}

my_df <- pivot_longer(as.data.frame(this_Counts_R2C), 1:6, names_to = "Sample",values_to = "CPM")
my_df$Sample <- factor(my_df$Sample, levels = c("WT_R1","WT_R2","WT_R3","R2C_R1","R2C_R2","R2C_R3"))
ggplot(my_df , aes(x = Sample, y = CPM,)) + geom_boxplot() + scale_y_continuous(trans='log2', limits = c(1, 16500)) + theme_bw()

```

```{r}

my_df <- pivot_longer(as.data.frame(this_Counts_R26C), 1:6, names_to = "Sample",values_to = "CPM")
my_df$Sample <- factor(my_df$Sample, levels = c("WT_R1","WT_R2","WT_R3","R26C_R1","R26C_R2","R26C_R3"))
ggplot(my_df , aes(x = Sample, y = CPM,)) + geom_boxplot() + scale_y_continuous(trans='log2', limits = c(1, 16500)) + theme_bw()

```

### Figure S4F

- Volcano plots of C+R

```{r}
DE_table_WTR2_CR <- read.table("GEO/final files/CR_HAatH3K27me3_R2C_minus_WT_DEG.tsv", header = T)
DE_table_WTR26_CR <- read.table("GEO/final files/CR_HAatH3K27me3_R26C_minus_WT_DEG.tsv", header = T)

EnhancedVolcano(DE_table_WTR2_CR,
    labSize=2,
    title = paste("HA signal at K27me3 SEACR peaks \n in WT vs R2C"),
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.05,
    subtitle ="",
    ylab = bquote(~-Log[10]~italic(padj)),
    legendLabels = c('NS', expression(Log[2]~FC),
    'padj', expression(padj~and~log[2]~FC)))

EnhancedVolcano(DE_table_WTR26_CR,
    labSize=2,
    title = paste("HA signal at K27me3 SEACR peaks \n in WT vs R62C"),
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.05,
    subtitle ="",
    ylab = bquote(~-Log[10]~italic(padj)),
    legendLabels = c('NS', expression(Log[2]~FC),
    'padj', expression(padj~and~log[2]~FC)))

```




